Modeling defoliation as a proxy for tree health: A case study using machine-learning and hyperspectral remote sensing data

### 2019-06-01 analysis/paper/Schratz_hyperspectral.tex
restructure
major refactoring of manuscript

@@

	\author[FSU]{Jannes Muenchow}
	\author[NEIKER]{Eugenia Iturritxa}
	%\author[TUDO]{Jakob Richter}
	\author[FSU]{Alexander Brenning}

	\address[FSU]{Department of Geography, GIScience group, Grietgasse 6, 07743, Jena, Germany}
	%\address[NEIKER]{NEIKER, Granja Modelo –Arkaute, Apdo. 46, 01080 Vitoria-Gasteiz, Arab, Spain}
	%\address[TUDO]{Department of Statistics, TU Dortmund University, Germany}

@@

	\begin{keyword}
		hyperspectral imagery \sep forest health \sep machine-learning \sep variable importance \sep model comparison
	\end{keyword}

@@

% Explain how remote sensing is used in forestry (potential to map forest health)
\noindent Data retrieved from remote sensing satellites is successfully used in forestry to monitor temporal changes across large areas \citep{martinezdelcastilloEvaluationForestCover2015,sextonModelPropagationUncertainty2015}.
The use of \ac{SAR} techniques enables scientists to estimate \ac{AGB} \citep{luSurveyRemoteSensingbased2016, sinhaReviewRadarRemote2015}.
Forest health is commonly assesed using optical data from multi-/hyperspectral satellites by applying either temporal change detections \citep{zhangRemoteSensingSeasonal2016} or by using vegetation indices to monitor the current state \citep{townsendGeneralLandsatModel2012}.
With the recent success story of machine-learning methods in the field of remote sensing, modeling techniques such as \ac{RF} are frequently used to model relationships of possible triggers to forest health \citep{belgiuRandomForestRemote2016, laryMachineLearningGeosciences2016, michezClassificationRiparianForest2016}.

With a robust model, predictions to large areas can be made, providing valuable information about the health condition of forest stands.
One approach to model forest health is to extract information from spectral information of affected and unaffected trees \citep{lelongEvaluationOilPalmFungal2010}.
Vegetation indices have shown the potential to provide valuable information to increase the predictive accuracy of forest pathogens \citep{jiangSatellitederivedVegetationIndices2014, adamczykRededgeVegetationIndices2015}.

However, the amount of possible (vegetation-)indices that can be calculated is often limited due to a low spectral resolution of freely available data from optical multispectral sensors (e.g. Sentinel-2).
Also, there is currently no freely available data from hyperspectral sensors that could be used for such studies (after the decommission of the EO-1 Hyperion satellite in January 2017).
To assess forest health on a tree scale, the spatial resolution of the data should be below 5 m as otherwise the value of a pixel may contain information from multiple trees and possibly even bare-ground information.
With this limitation, data from sateillites with a coarser resolution (e.g. Sentinel-2) can be used for prediction tasks but not for training purposes of algorithms on a tree level.

% Why are we mapping defoliation?

In this study we will use hyperspectral data with a spatial resolution of one meter and 126 spectral bands to model the health status of Monterey Pine (\textit{Pinus radiata}) plantations in northern Spain (\autoref{fig:study_area}).
The trees in the study area suffer from infections of invasive pathogens such as \textit{Diplodia sapinea}, \textit{Fusarium circinatum}, \textit{Armillaria mellea} or \textit{Heterobasidion annosum} leading to a spread of cankers or defoliation \citep{mesanzaNativeRhizobacteriaBiocontrol2016, iturritxaBiocontrolFusariumCircinatum2017}.
In-situ measurements of defoliation on a tree level (as a proxy for tree health) have been collected to serve as the response variable.
The fungi are assumed to infect the trees through open wounds, possibly caused by previous hail damage \citep{Iturritxa2014}.
The dieback of these trees, which are mainly used as timber, causes high economic damages \citep{ganleyGlobalClimaticRisk2009}.

We used state-of-the-art machine-learning techniques in combination with high resolution remote sensing information to model forest health on a tree level.
Furthermore, we aimed to reduce the amount of predictors using feature selection to simplify the spatial prediction of the fitted model.

\noindent Specifically the following objectives were addressed:

\begin{itemize}
	\item Performance comparison of multiple algorithms modeling defoliation of \textit{Pinus radiata} trees using highly-correlated features
	\item Exploration of the most important variables
	\item Spatial prediction of \textit{Pinus radiata} defoliation
\end{itemize}

\section{Data and study area}

@@

\noindent The \textit{Pinus radiata} plots of this study, namely \textit{Laukiz 1}, \textit{Laukiz 2}, \textit{Luiando} and \textit{Oiartzun}, are located in the northern part of the Basque Country (\autoref{fig:study_area}).
\textit{Oiartzun} has the most observations (n = 529) while \textit{Laukiz 2} has the largest area size (1.44 ha).
All plots besides \textit{Luiando} are located nearby the coast (\autoref{fig:study_area}).
In total 1750 observations are available (\textit{Laukiz 1} = 479, \textit{Laukiz 2} = 451, \textit{Luiando} = 291, \textit{Oiartzun} = 529).
The data was surveyed in September 2016.

\begin{figure} [t!]
	\begin{center}
		\makebox[\textwidth]{\includegraphics[width=\textwidth] {study_area_hyperspectral.pdf}}
		\caption{Information about the plot locations, the area of hyperspectral coverage and the number of trees per plot.}
		\label{fig:study_area}
	\end{center}
\end{figure}

@@

\noindent The airborne hyperspectral data was acquired during two flight campaigns on September 28th and October 5th 2016, both around 12 am.
The images were taken using a AISAEAGLE-II sensor.
All preprocessing steps (geometric, radiometric, atmospheric) have been conducted by the \ac{ICGC}.
The first four bands are corrupted, leaving 122 bands with valid information.
Additional metadata information is available in Table 1:

@@

\noindent For all analysis steps we used the open-source statistical programming language R \citep{R_core}.
The algorithm implementations of the following packages have been used: \textit{xgboost} \citep{chenXGBoostScalableTree2016} (\textit{xgboost}), \textit{kernlab} \citep{kernlab} (Support Vector Machine) and \textit{glmnet} \citep{glmnet} (Ridge Regression).
We used the R package \textit{mlr} for all modeling related steps.
It provides a standardized interface for a wide variety of statistical and machine-learning models in R simplifying essential modeling tasks such as hyperparameter tuning, model performance evaluation and parallelization \citep{bischlMlrMachineLearning2016}.
We provide the complete code and required data as a Mendeley dataset to make this work fully reproducible [STILL TO COME].

\noindent To use the full information from the hyperspectral data, we calculated all possible vegetation indices that are available in the \textit{hsdar} package (90 in total) and all possible \ac{NRI} combinations.
We were interested if NRIs of arbitrary band combinations will have a substantial effect on the predictive power of the fitted model.
The NRIs were calculated using the following formula:

@@

@@

\noindent Three algorithms (\textit{xgboost}, \textit{\ac{SVM}} and \textit{\ac{RR}}) were benchmarked on their predictive performance.
Besides the well-known \ac{SVM} algorithm \citep{Vapnik1998}, we also used \textit{xgboost} which is ensemble method relying on the idea of tree boosting that gained a lot of attention in recent years \citep{chenXGBoostScalableTree2016}.
We added penalized L2 (Ridge) regression to the portfolio due to its ability to handle highly correlated covariates \citep{hoerlRidgeRegressionBiased1970}.
One of the most popular machine-learning algorithm, Random Forest \citep{Breiman2001}, was not considered for this study: Due to the high number of variables, model fitting times in the range hours for a single model fit were not practicable for this work.
These high fitting times are caused by hyperparameter \texttt{mtry} which scales with the number of variables \citep{Probst2018b}.
After the selection of the best model, we checked if the winning algorithm can achieve a similar performance when using only the most important variables compared to using all variables.
A successful feature selection simplifies the spatial prediction task because the prediction dataset needs to consist of less variables.
Furthermore, model complexity and fitting times are reduced.

@@

\begin{figure} [t!]
	\begin{center}
		\makebox[\textwidth]{\includegraphics[width=1\textwidth] {spcv_folds_demon_plots.pdf}}
		\caption{Fold 1 of the spatial partitioning using k-means clustering for \textit{Laukiz 1}, \textit{Laukiz 2}, \textit{Luiando} and \textit{Oiartzun}.}
		\label{fig:spcv}
	\end{center}
\end{figure}

@@

\noindent To tune the hyperparameters of the algorithms, we used \ac{SMBO} via the R package \textit{mlrMBO} \citep{mlrMBO}.
This Bayesian approach first composes \textit{n} randomly chosen hyperparameter settings out of a user defined search space.
After these \textit{n} tries have been evaluated, a new hyperparameter setting, which is going to be evaluated next, is proposed based on a fitted regression model.
The regression model estimates the performance of the machine-learning method for unknown hyperparameter settings.
Using these estimates, a new promising hyperparameter setting is proposed to be evaluated next.
This strategy continues until a termination criterion, defined by the user, is reached \citep{Hutter2011, Jones1998}.

In this work we used an initial design of 30 randomly composed hyperparameter settings and a termination criterion of 20 iterations, resulting in a total budget of 50 evaluated hyperparameter settings per fold.
The advantage of this tuning approach is that it substantially reduces the tuning budget which is needed to find a setting close to the global minimum compared to methods that do not use information from previous runs, such as random search or grid search \citep{Bergstra2012}.

\noindent To find indices that contributed most to model performance, we used the internal variable importance measure of the \textit{xgboost} algorithm.
This score is calculated by taking the contribution of each feature for each tree in the fitted model.
The higher the score of a variable, the more important it is for the fitted model when making predictions \citep{chenXGBoostScalableTree2016}.
The variable importance measure is automatically computed during model fit.
In contrast to other approaches such as permutation-based ones, the \textit{xgboost} score is composed out of three parts that contribute to the overall importance \citep{chenXGBoostScalableTree2016}:

\begin{itemize}
	\item Gain: The relative contribution of the feature to the model
	\item Cover metric: How often a feature was selected to be the deciding feature in a tree for a specific observation
	\item Frequency How often a feature occurs in all trees of the model
\end{itemize}

\noindent The \textit{Gain} features is the most important one among the three.
All measures sum up to one \citep{chenXGBoostScalableTree2016}.

% partial dependence plots??

@@

% defoliation boxplots
\begin{figure} [t!]
	\begin{center}
		\makebox[\textwidth]{\includegraphics[width=0.7\textwidth] {boxplot_defol.pdf}}
		\caption{Descriptive statistics of the response variable \textit{defoliation}.}
		\label{fig:defol_boxplots}
	\end{center}
\end{figure}

\noindent \textit{Oiartzun} shows the highest defoliation ($\bar{x} = 69.22 \%$) among the plots while \textit{Laukiz 2} is the healthiest ($\bar{x} = 13.54 \%$) (\autoref{fig:defol_boxplots}).
All plots besides \textit{Luiando} show an evenly distributed level of defoliation across the entire plot.

\noindent The high degree of defoliation for \textit{Luiando} and \textit{Oiartzun} is also visible in the spectral signatures of the plots (\autoref{fig:spectral_signatures}).
Both plots show lower mean reflectance values around the wavelength range 800 nm - 1000 nm compared to \textit{Laukiz 1} and \textit{Laukiz 2}.
\textit{Oiartzun} is almost completely missing the reflectance drop at around 815 nm that is visible for all other plots but instead shows a higher magnitude for the reflectance increase at around 920 nm.

\noindent \textit{Laukiz 2} shows a mean tree density of 61.59 m (\autoref{fig:plot-characteristics}) while all other plots have a higher density (34.64 m (Laukiz 1), 33.01 m (Luiando), 34.96 m (Oiartzun)) (\autoref{fig:plot-characteristics}).

@@

% Performance estimates
\begin{table}[t!]
\centering
\caption[t]{Spatial block \ac{CV} performances of \textit{RR}, \textit{SVM} and \textit{xgboost} using \ac{RMSE} as the error measure.
	Mean and standard deviation are shown.}
\begingroup\footnotesize
\begin{tabular}{llll}
	RR            & SVM           & xgboost       & xgboost (7 variables) \\
	\hline
	59.10 (22.71) & 36.23 (15.73) & 33.26 (16.61) & 29.59   (16.09)       \\
	\bottomrule
\end{tabular}
\endgroup
\label{tab:model_comparison}
\end{table}

\begin{table}[b!]
\centering
\caption[t]{Predictive performance of \textit{xgboost} using all observations and all variables (All Observations/all variables), all observations and the seven most important variables only (All Observations/7 variables) and observations from specific plots only (Plot level observation/all variables) with \ac{RMSE} as the error measure. The performance estimates for "All Observations" correspond to the fold for which the respective plot was serving as the test set (block CV). Column "Plot level observations", shows the mean RMSE estimates at the repetition level of a five-fold five-time repeated spatial CV, scored by using data of the respective plot only.}
\begingroup\footnotesize
\begin{tabular}{llll}
	\\
	Plot/Data & \specialcell{All Observations/ \\{all variables (Block CV)}} & \specialcell{All Observations/         \\ {7 variables (Block CV)}} & \specialcell{Plot level observations/\\{all variables (SpCV)}} \\
	\hline
	Laukiz 1  & 22.03                                                                              & 21.47                          & 19.18 \\
	Laukiz 2  & 51.75                                                                              & 49.94                          & 17.24 \\
	Luiando   & 13.20                                                                              & 15.37                          & 8.30  \\
	Oiartzun  & 32.97                                                                              & 17.62                          & 14.40 \\
	\bottomrule
\end{tabular}
\endgroup
\label{tab:supermodel_performance}
\end{table}

@@

\noindent The \textit{xgboost} algorithm showed the lowest error (33.26 RMSE) when benchmarking the learners on the complete dataset of all plots (\autoref{tab:model_comparison}).
While the \textit{SVM} performance was only slightly worse (36.23 RMSE), \textit{RR} showed a substantially worse performance than \textit{xgboost} (59.10 RMSE).

\subsubsection{Single models vs. super model}

\noindent Comparing the mean predictive performance of models fitted at the plot level against the performance of the model that was fitted using all data (super model), the plot-level models showed a better performance in all cases (\autoref{tab:supermodel_performance}).
The highest difference between both datasets occurred for plot \textit{Laukiz 2} with a difference of 34.51 RMSE.

Using only the seven most important variables (\autoref{fig:var-imp}) for the super model showed small increases in performance for \textit{Laukiz 1} and \textit{Laukiz 2}, a small decrease for \textit{Luiando} and almost a reduction of 50\% of the error for \textit{Oiartzun} (32.97 vs. 17.62 RMSE) (\autoref{tab:supermodel_performance}).

\subsubsection{RMSE vs. plot characteristics}

\noindent An increase of the error rate was observed with an increase of descriptive plot measures such as mean point density and the coefficient of variation (based on the response variable \textit{defoliation}) (\autoref{fig:plot-characteristics}).

\begin{figure} [t!]
	\begin{center}
		\makebox[\textwidth]{\includegraphics[width=\textwidth] {plot-characteristics.pdf}}
		\caption{RMSE vs. mean point density and coefficient of variation (defoliation).}
		\label{fig:plot-characteristics}
	\end{center}
\end{figure}

\noindent The seven most important features of the super model in this study were vegetation indices with the \textit{EVI} \citep{hueteComparisonVegetationIndices1997} being the most important one (\autoref{fig:var-imp}).

\begin{equation}
	EVI = 2.5*\frac{R_{800}-R_{670}}{R_{800}-(6*R_{670})-(7.5*R_{475})+1)}
\end{equation}

\noindent where $R$ = Reflectance at the respective wavelength.

\noindent Vegetation index \textit{GDVI} appears three times among the first seven most important features (\autoref{fig:var-imp}) with different \texttt{n} values.
This is because it was computed four times, with \texttt{n} ranging from 1 - 4 \citep{wuEstimatingChlorophyllContent2008}:

\begin{equation}
	GDVI = \frac{R_{800}^n-R_{680}^n}{R_{800}^n+R_{680}^n}
\end{equation}

\noindent The seven most important features (\textit{EVI}, \textit{GDVI\_4}, \textit{D1}, \textit{GDVI\_3}, \textit{GDVI\_2}, \textit{mNDVI} and \textit{mSR}) showed a substantial difference in the importance score compared to all following variables (\autoref{fig:var-imp}).

\begin{figure} [b!]
	\begin{center}
		\makebox[\textwidth]{\includegraphics[width = 1\textwidth] {var-imp-xgboost.pdf}}
		\caption{The 30 most important variables as estimated by the internal variable importance measure of the \textit{xgboost} algorithm. The higher the score, the more important the feature.
			"bf2" notes that a buffer of 2 meter was used to extract the variable information to the tree observation. "NRI" means that a normalized ratio index with the subsequent bands was calculated. Features without "NRI" prefix are vegetation indices, e.g. "bf2\_EVI".}. %An overview of the vegetation indices can be found \href{https://rdrr.io/cran/hsdar/man/vegindex.html}{here}.}
		\label{fig:var-imp}
	\end{center}
\end{figure}

\begin{table}[t!]
\centering
\caption[t]{Formulas of the five most important vegetation indices of the super model. $R$ = Reflectance at wavelength, $D$ = First derivation of reflectance value at wavelength.}
\begingroup\footnotesize
\begin{tabular}{llll}
	\\
	Acronym & Name                      & Formula                                                            & Reference                                                 \\
	\hline
	EVI     & Enhanced vegetation index & $2.5*\frac{R_{800}-R_{670}}{R_{800}-(6*R_{670})-(7.5*R_{475})+1)}$ & \cite{hueteComparisonVegetationIndices1997}               \\
	GDVI    & Generalized DVI*          & $\frac{R_{800}^n-R_{680}^n}{R_{800}^n+R_{680}^n}$                  & \cite{wuEstimatingChlorophyllContent2008}                 \\
	D1      & Derivative Index          & $\frac{D_{730}}{D_{706}}$                                          & \cite{zarco-tejadaSteadystateChlorophyllFluorescence2003} \\
	mNDVI   & Normalized DVI*           & $  \frac{R_{800}-R_{680}}{(R_{800}+R_{680}-2*R_{445}}$             & \cite{simsRelationshipsLeafPigment2002}                   \\
	mSR     & Simple Ratio Index        & $\frac{R_{800}-R_{445}}{R_{680}-R_{445}}$                          & \cite{simsRelationshipsLeafPigment2002}                   \\
	\bottomrule
	\multicolumn{2}{l}{\small{* Difference Vegetation Index}}
\end{tabular}
\endgroup
\label{tab:most_imp_vars}
\end{table}

\noindent The best NRI scored rank eight (band 112 and band 62).
All further places up to rank 30 were occupied by NRIs.

\begin{figure} [b!]
	\begin{center}
		\makebox[\textwidth]{\includegraphics[width = 1\textwidth] {predictions_demon_plots.pdf}}
		\caption{Spatially predicted defoliation (in \%) from \textit{xgboost} of \textit{Laukiz 1}, \textit{Laukiz 2}, \textit{Luiando} and \textit{Oiartzun}.}
		\label{fig:pred_demon_plots}
	\end{center}
\end{figure}

\begin{figure} [t!]
	\begin{center}
		\makebox[\textwidth]{\includegraphics[width = 0.6\textwidth] {predictions_demon_plots_hists.pdf}}
		\caption{Histograms of predicted defoliation (in \%) from \textit{xgboost} of \textit{Laukiz 1}, \textit{Laukiz 2}, \textit{Luiando} and \textit{Oiartzun}.}
		\label{fig:pred_demon_plots_hists}
	\end{center}
\end{figure}

\noindent The plots with a higher mean defoliation (\textit{Luiando} and \textit{Oiartzun}) showed a good visual separability compared to the healthier plots \textit{Laukiz 1} and \textit{Laukiz 2} (\autoref{fig:pred_demon_plots}).
All predicted values ranged between 30 \% and 80 \% defoliation with \textit{Luiando} showing the smallest variance (\autoref{fig:pred_demon_plots_hists}).

The high error of the super model for \textit{Laukiz 2} (49.94 RMSE) is also visible in the respective histogram as most predictions range between a defoliation of 40 \% - 50 \% (\autoref{fig:pred_demon_plots_hists}) while in fact most trees of \textit{Laukiz 2} show an actual defoliation of around 20 \% (\autoref{fig:defol_boxplots}).

For the less defoliated plots \textit{Laukiz 1} and \textit{Laukiz 2} a subtle level of separation between trees and bare ground is visible (\autoref{fig:pred_demon_plots}).
This is not the case for the other two higher defoliated plots for which bare ground and defoliated predictions are mainly in the same value range (around 60 \% - 80 \%).

@@

\noindent The relatively large difference in performance between \textit{RR} (59.10 RMSE) and the machine-learning models (36.23 and 33.26 RMSE) is remarkable.
\textit{RR} has shown promising results in other studies when many highly-correlated predictors were involved \citep{hernandezUsingRidgeRegression2015, imaniRidgeRegressionbasedFeature2015}.
However, in this study, \textit{RR} was not able to achieve a sufficient performance score compared to \textit{SVM} and \textit{xgboost} even though its hyperparameter $\lambda$ was properly tuned using SMBO (\autoref{meth:smbo}).

While \textit{xgboost} showed a slightly better performance than \textit{SVM}, the latter has the advantage of only having two hyperparameters that need to be tuned.
This results in a shorter runtime.
Nevertheless, \textit{xgboost} showed the best performance and was subsequently selected to fit the models on the plot level and for the spatial prediction.

An important point that needs to be considered when interpreting the performance results is that we only related defoliation to indices derived from remote sensing data.
Possible other local variables that could help in predicting defoliation were not considered.
One example here is tree age: The older a tree the more vulnerable it may be to pathogen infections causing defoliation.
However, such predictors would not be available for a spatial prediction scenario and one of the main goals of this study is to relate defoliation to variables that are available on a larger scale (e.g. remote sensing indices).

\subsubsection{Single models vs. super model}

\noindent It is expected that models that were trained using observations from the respective plot only only achieve a better performance than the super model which was trained on observations from multiple plots (\autoref{tab:supermodel_performance}).
The low performance on \textit{Laukiz 2} for the super model is most likely due to the difference of this plot to all others: The fitted model on \textit{Laukiz 1}, \textit{Luiando} and \textit{Oiartzun} is not capable of reaching a good performance when \textit{Laukiz 2} is the evaluation dataset.
This is not surprising as \textit{Laukiz 2} shows substantially different plot characteristics compared to all others plots in terms of the distribution of the response variable \textit{defoliation} (\autoref{fig:defol_boxplots}) and the mean point density of trees (\autoref{fig:plot-characteristics}).

The low error for Luiando (8.30 RMSE) for the plot model validates the approach of relating defoliation to vegetation indices and NRIs.
The overall error of the super model (33.26 RSME) is expected to decrease if more plots would be available for training.
However, even if the fitted model would include at least one instance of every plot showing unique characteristics (e.g. here \textit{Laukiz 2} is substantially different to the others), the overall error would not become smaller than the error achieved when using data from the respective plot only.

An interesting find is that the model with only seven variables shows a better overall performance than the model with all 7471 variables (\autoref{tab:supermodel_performance}).
This leads to the conclusion that adding as many variables as possible to a model will not necessarily improve its performance but instead add noise to the model.
Too much information can be problematic for a model as it will have a hard time distinguishing between noise and important information in the variables \citep{liFeatureSelectionData2017, guyonIntroductionVariableFeature2003}.
However, to find the most important variables in the first place and to check for the performance difference, a model with all variables needs to fitted first.
Using a model with only a few predictors does not only simplify prediction tasks but also reduces runtime for hyperparameter tuning and performance estimation \citep{liuComputationalMethodsFeature2007}.

@@

\noindent There are some downsides using the internal variable importance approach of \textit{xgboost}: Due to the contribution of three different parts to the overall importance score it is complicated to understand why a specific feature was selected.
Furthermore the importance calculation approach is only valid for this algorithm and cannot be compared to others.
Nevertheless, as we only relied on the variable importance for this specific algorithm, using the internal \textit{xgboost} approach was sufficient for this work.

It is expected that vegetation indices are most important for the model as these are most sensitive to changes in vegetation health \citep{croftApplicabilityEmpiricalVegetation2014}.
Even though we are not directly looking at vegetation health but using the level of defoliation as a proxy for tree health, vegetation indices were most important for the fitted model of this study (\autoref{fig:var-imp}).

\noindent Vegetation indices can help assessing defoliation in two ways:

\begin{itemize}
	\item Trees that show a high level of defoliation do also reflect their bad health status through the remaining foliation.
	\item Defoliated trees have more influence of bare ground information in their pixel values and will therefore be classified as defoliated by the model.
\end{itemize}

\noindent Even though no NRI made it among the most important variables in this study (\autoref{fig:var-imp}) (stating that the first seven of this study are the most important ones), it is notable that all ranks from 8 - 30 are occupied by NRI (\autoref{fig:var-imp}).
However, their relative importance was very small compared to the first seven ranks.

Restricting the important indices of this study to the first seven ranks can be seen critical as we only based the selection on a visual inspection of the variable importance results (\autoref{fig:var-imp}).
The decision to make a cut between rank seven and eight was based on a combination of two facts:

\begin{itemize}
	\item Using only vegetation indices is easier for large scale predictions using satellites like Sentinel-2 (most NRIs cannot be used with it because the spectral bands do not exist).
	\item The drop in the importance score of the variable importance results (\autoref{fig:var-imp}).
\end{itemize}

\noindent However, based on these two points, we could also have made the cut between rank five and 6 but including the two vegetation indices at rank six and seven will eventually improve the model and does not increase runtime.

>>>>

% https://tex.stackexchange.com/questions/165115/getting-not-defining-perthousnad-and-not-defining-micro-when-compiling-beamer
\usepackage{textcomp}

@@

	\author[FSU]{Jannes Muenchow}
	\author[NEIKER]{Eugenia Iturritxa}
	\author[LMU]{Bernd Bischl}
	\author[FSU]{Alexander Brenning}

	\address[FSU]{Department of Geography, GIScience group, Grietgasse 6, 07743, Jena, Germany}
	\address[NEIKER]{NEIKER, Granja Modelo –Arkaute, Apdo. 46, 01080 Vitoria-Gasteiz, Arab, Spain}
	\address[LMU]{Department of Statistics, Chair for computational Statistics, Ludwig-Maximilian University Munich, Germany}

@@

	\begin{keyword}
		hyperspectral imagery \sep forest health modeling \sep machine-learning \sep feature-selection  \sep model comparison
	\end{keyword}

@@

% Explain how remote sensing is used in forestry (potential to map forest health)
% Link remote sensing and machine-learning
The use of machine-learning algorithms for analyzing remote sensing data has seen a huge increase in the last decade.
Ever since the launch of the first Sentinel satellite in 2014 the amount of freely available imagery data has increased enormously (REF).
At the same time, the implementation of algorithms (both statistical and machine-learning) has been greatly simplified in the open-source community.
Scientists can nowadays relively easily feed large amounts of data into complex algorithms in a semi-automated way.
These facts combined with a high temporal and spatial coverage of input data enable the possibility to analyze temporal changes of the environment on a large scale.

% link to forest health analysis and show exemplary studies
The monitoring of forest ecosystems is one exemplary field that profits from these new possibilities.
Specifically the monitoring of forest health got simplified with the ability of analyzing large scale data in efficient algorithms that are able to predict to unknown areas.
For this task, usually optical data from multi-/hyperspectral satellites is used.
Various indices are calculated from the data with the aim of carrying out temporal change detections \citep{zhang2016} or analyses about the forest health status for a specific date \citep{townsend2012}.
% TODO: check if refs match context
Vegetation indices have shown the potential to provide valuable information for analyzing triggers to forest health \citep{jiang2014, adamczyk2015}.
More and more studies focus on using machine-learning algorithms such as \ac{RF} \citep{belgiu2016, lary2016, michez2016}.
Overall, the idea is to fit a robust model which can be used to make predictions to (large) unknown areas, providing valuable information about the health condition of forest stands without the need for ground-truth data in these regions.

% now mention why this study is important
However, we have the impression that the amount of (vegetation-)indices used in forest health studies could be higher, possibly leading to more valuable results.
Possible reasons include the low spectral resolution of optical multispectral sensors, the lack of freely available hyperspectral data or the lack of knowledge about existing indices.
% TODO: Ref for high-dim
Another more applied issue is the fact of high-dimensionality which is introduced by the calculation of hundreds/thousands of indices.
% TODO: Ref model interpr
Even though ML algorithms are capable of handling possibly highly-correlated input variables, model runtime increases substantially and the interpretation of results becomes more complicated.
Also, in recent years the awareness of conducting proper tuning of model hyperparameters has increased in applied modelling fields.
In this study we addressed these challenges by focusing on the following research questions:

\begin{itemize}
	\item How to handle and analyze high-dimensional data including model optimization while accounting for spatial characteristics?
	\item Does the calculation of arbitrary indices add any value to the analysis or is the band information from the sensor already sufficient?
	\item How can interpretability of the model be maintained while at the same time using all possible input data?
\end{itemize}

\noindent State-of-the-art machine-learning techniques were compared using three feature sets across different feature-selection approaches including a relatively novel approach of ensemble filters.
To model the numeric response variable defoliation at trees model-based optimization was applied for hyperparameter tuning and a spatial block \ac{CV} was used to account for spatial autocorrelation in the data.

\section{Data and study area}
\noindent Airborne hyperspectral data with a spatial resolution of one meter and 126 spectral bands was available for four Monterey Pine (\textit{Pinus radiata}) plantations in northern Spain (\autoref{fig:study_area}).
The trees in the study area plots suffer from infections of invasive pathogens such as \textit{Diplodia sapinea}, \textit{Fusarium circinatum}, \textit{Armillaria mellea} or \textit{Heterobasidion annosum} leading to a spread of cankers or defoliation \citep{mesanza2016, iturritxa2017}.
In-situ measurements of defoliation at trees (as a proxy for tree health) were collected to serve as the response variable \textit{defoliation} spanning a range from 0\% - 100\%.
The fungi are assumed to infect the trees through open wounds, possibly caused by previous hail damage \citep{Iturritxa2014}.
The dieback of these trees, which are mainly used as timber, causes high economic damages \citep{Ganley2009}.

@@

\noindent The \textit{Pinus radiata} plots of this study, namely \textit{Laukiz 1}, \textit{Laukiz 2}, \textit{Luiando} and \textit{Oiartzun}, are located in the northern part of the Basque Country (\autoref{fig:study_area}).
\textit{Oiartzun} has the most observations (n = 529) while \textit{Laukiz 2} features the largest area size (1.44 ha).
All plots besides \textit{Luiando} are located nearby the coast (\autoref{fig:study_area}).
In total 1750 observations are available (\textit{Laukiz 1} = 479, \textit{Laukiz 2} = 451, \textit{Luiando} = 291, \textit{Oiartzun} = 529).
The data was surveyed in September 2016.

%\begin{figure} [t!]
%\begin{center}
%\makebox[\textwidth]{\includegraphics[width=\textwidth] {study_area_hyperspectral.pdf}}
%\caption{Information about the plot locations, the area of hyperspectral coverage and the number of trees per plot.}
%\label{fig:study_area}
%\end{center}
%\end{figure}

@@

\noindent The airborne hyperspectral data was acquired during two flight campaigns on September 28th and October 5th 2016, both around 12 am.
The images were taken using a AISAEAGLE-II sensor.
All preprocessing steps (geometric, radiometric, atmospheric) have been conducted by the \ac{ICGC}.
The first four bands are corrupted, leaving 122 bands with valid information.
Additional metadata information is available in \autoref{tab:hyperparameter_limits}.

@@

\subsection{Derivation of indices}
% TODO: Mention different feature sets
\noindent To use the full information from the hyperspectral data, we calculated all possible vegetation indices supported by the \textit{hsdar} package (90 in total) as well as all possible \ac{NRI} combinations.
NRIs were calculated using the following formula:

@@

\subsection{Feature selection methods}

% TODO: List filter methods
% TODO: Explain the idea of ensemble filters

@@

\noindent Three algorithms (\textit{xgboost}, \textit{\ac{SVM}} and \textit{\ac{RR}}) were benchmarked on their predictive performance.
Besides the well-known \ac{SVM} algorithm \citep{vapnik1998}, we also used \textit{xgboost} which is ensemble method relying on the idea of tree boosting that gained a lot of attention in recent years \citep{chen2016}.
We added penalized L2 (Ridge) regression to the portfolio due to its ability to handle highly correlated covariates \citep{hoerl1970}.
One of the most popular machine-learning algorithm, Random Forest \citep{breiman2001}, was not considered for this study: Due to the high number of variables, model fitting times in the range hours for a single model fit were not practicable for this work.
These high fitting times are caused by hyperparameter \texttt{mtry} which scales with the number of variables \citep{probst2018}.
After the selection of the best model, we checked if the winning algorithm can achieve a similar performance when using only the most important variables compared to using all variables.
A successful feature selection simplifies the spatial prediction task because the prediction dataset needs to consist of less variables.
Furthermore, model complexity and fitting times are reduced.

@@

%\begin{figure} [t!]
%\begin{center}
%\makebox[\textwidth]{\includegraphics[width=1\textwidth] {spcv_folds_demon_plots.pdf}}
%\caption{Fold 1 of the spatial partitioning using k-means clustering for \textit{Laukiz 1}, \textit{Laukiz 2}, \textit{Luiando} and \textit{Oiartzun}.}
%\label{fig:spcv}
%\end{center}
%\end{figure}

@@

\noindent To tune the hyperparameters of the algorithms, we used \ac{SMBO} via the R package \textit{mlrMBO} \citep{mlrmbo}.
This Bayesian approach first composes \textit{n} randomly chosen hyperparameter settings out of a user defined search space.
After these \textit{n} tries have been evaluated, a new hyperparameter setting, which is going to be evaluated next, is proposed based on a fitted regression model.
The regression model estimates the performance of the machine-learning method for unknown hyperparameter settings.
Using these estimates, a new promising hyperparameter setting is proposed to be evaluated next.
This strategy continues until a termination criterion, defined by the user, is reached \citep{hutter2011, jones1998}.

In this work we used an initial design of 30 randomly composed hyperparameter settings and a termination criterion of 20 iterations, resulting in a total budget of 50 evaluated hyperparameter settings per fold.
The advantage of this tuning approach is that it substantially reduces the tuning budget which is needed to find a setting close to the global minimum compared to methods that do not use information from previous runs, such as random search or grid search \citep{bergstra2012}.

\noindent The complete study was done using the open-source statistical programming language R \citep{rcoreteam2018}.
The algorithm implementations of the following packages have been used: \textit{xgboost} \citep{chen2016} (\textit{xgboost}), \textit{e1071} \citep{e1071} (Support Vector Machine) and \textit{glmnet} \citep{glmnet} (Ridge Regression).
The filter methods of the following packages were used: \textit{praznik}, \textit{fselectorrcpp}.
The R package \textit{mlr} \citep{mlr} was used for all modeling related steps and \textit{drake} for setting up the workflow of the project.
This study is available as a research compendium on Zenodo (\url{10.5281/zenodo.2635403}).

@@

%% defoliation boxplots
%\begin{figure} [t!]
%\begin{center}
%\makebox[\textwidth]{\includegraphics[width=0.7\textwidth] {boxplot_defol.pdf}}
%\caption{Descriptive statistics of the response variable \textit{defoliation}.}
%\label{fig:defol_boxplots}
%\end{center}
%\end{figure}

%\noindent \textit{Oiartzun} shows the highest defoliation ($\bar{x} = 69.22 \%$) among the plots while \textit{Laukiz 2} is the healthiest ($\bar{x} = 13.54 \%$) (\autoref{fig:defol_boxplots}).
%All plots besides \textit{Luiando} show an evenly distributed level of defoliation across the entire plot.

%\noindent The high degree of defoliation for \textit{Luiando} and \textit{Oiartzun} is also visible in the spectral signatures of the plots (\autoref{fig:spectral_signatures}).
%Both plots show lower mean reflectance values around the wavelength range 800 nm - 1000 nm compared to \textit{Laukiz 1} and \textit{Laukiz 2}.
%\textit{Oiartzun} is almost completely missing the reflectance drop at around 815 nm that is visible for all other plots but instead shows a higher magnitude for the reflectance increase at around 920 nm.

%\noindent \textit{Laukiz 2} shows a mean tree density of 61.59 m (\autoref{fig:plot-characteristics}) while all other plots have a higher density (34.64 m (Laukiz 1), 33.01 m (Luiando), 34.96 m (Oiartzun)) (\autoref{fig:plot-characteristics}).

@@

@@

@@

@@
